---
title: "Antimicrobial Resistance Analysis of Human Non-typhoidal Salmonella, Malaysia, 2018–2024"
author: "Hazlienor Mohd Hatta"
subtitle: "R script for DrPH Thesis: Incidence, Serovar Dynamics and Forecasting of Human Non-typhoidal Salmonella Disease in Malaysia"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
editor: visual
---

# Introduction

This Quarto document contains the R workflow for the antimicrobial resistance (AMR) analysis conducted for the DrPH thesis titled **"Incidence, Serovar Dynamics and Forecasting of Human Non-typhoidal Salmonella Disease in Malaysia"** by **Hazlienor Mohd Hatta**.

The analysis uses human non-typhoidal *Salmonella* (NTS) surveillance data from Malaysia, 2018–2024. The workflow includes data import, cleaning of antimicrobial susceptibility testing (AST) results, classification of drug-resistant (DR) and multidrug-resistant (MDR) isolates, estimation of resistance proportions, and assessment of AMR patterns by year, serovar, state/region, specimen type, and age group.

> **Note:** This script is intended for reproducible analysis. No individual-level identifiable patient information should be shared in the public repository.

# Analysis workflow

The AMR analysis follows these steps:

1.  Load required R packages and import WHONET-derived AMR data.
2.  Clean AST variables and standardise susceptibility results.
3.  Define antimicrobial classes, drug resistance (DR), and multidrug resistance (MDR).
4.  Estimate overall resistance proportions by antibiotic.
5.  Assess temporal trends in resistance by year.
6.  Describe DR and MDR patterns by serovar.
7.  Examine geographical variation by state and region.
8.  Examine AMR patterns by specimen type and age group.
9.  Summarise common MDR resistance combinations.

# Setup

```{r setup, message=FALSE, warning=FALSE}
# Core data manipulation and plotting
library(tidyverse)
library(readxl)
library(ggplot2)

# Model summaries and confidence intervals
library(broom)
library(binom)

# Reordering labels within facets
library(tidytext)

# Optional: for regression tables if needed
# library(sjPlot)
```

# Data import

```{r import-data}
# Import AMR dataset.
# Update the file path if the data file is stored in a different folder.
AMR_raw <- read_excel("IMR Rfinal.xlsx", col_types = "text")

# Create working dataset
AMR <- AMR_raw
```

# Define antibiotic variables

```{r antibiotic-list}
# Antibiotic columns included in the AMR analysis.
# These columns should contain AST interpretations, usually S, I, or R.
ast_cols <- c(
  "AMP", "PIP", "TZP", "CTX", "CRO", "IPM", "MEM", "ETP", "AZM", "TCY",
  "CIP", "NOR", "PEF", "CXM", "FEP", "CFP", "CSL", "CAZ", "SXT", "CHL",
  "AMK", "GEN", "AMC", "SAM", "NET", "NIT"
)

# Check whether all specified antibiotic columns are available in the dataset
missing_ast_cols <- setdiff(ast_cols, names(AMR))
missing_ast_cols
```

# Data cleaning

```{r clean-data}
# Standardise AST results:
# - convert to uppercase
# - remove extra spaces
# - convert non-informative values to missing
AMR <- AMR %>%
  mutate(across(all_of(ast_cols), ~ {
    x <- as.character(.)
    x <- str_trim(toupper(x))
    x[x %in% c("", "NA", "N/A", "ND", ".", "-", "NULL")] <- NA_character_
    x
  })) %>%
  mutate(
    year = as.numeric(as.character(year))
  )

# Optional: inspect key variables
# glimpse(AMR)
# table(AMR$year, useNA = "ifany")
```

# Define antimicrobial classes, DR and MDR

```{r define-dr-mdr}
# Define resistance to antimicrobial classes.
# MDR is defined here as resistance to three or more antimicrobial classes.
# DR is defined as resistance to at least one antimicrobial class.
AMR <- AMR %>%
  mutate(
    beta_lactam      = if_any(c(AMP, PIP, TZP, AMC, SAM), ~ . == "R"),
    cephalosporin    = if_any(c(CTX, CRO, CAZ, FEP, CXM, CFP), ~ . == "R"),
    carbapenem       = if_any(c(IPM, MEM, ETP), ~ . == "R"),
    aminoglycoside   = if_any(c(AMK, GEN, NET), ~ . == "R"),
    fluoroquinolone  = if_any(c(CIP, NOR, PEF), ~ . == "R"),
    tetracycline     = TCY == "R",
    sulfonamide      = SXT == "R",
    amphenicol       = CHL == "R",
    macrolide        = AZM == "R",
    nitrofuran       = NIT == "R"
  ) %>%
  mutate(
    n_classes_resistant = rowSums(across(
      c(beta_lactam, cephalosporin, carbapenem, aminoglycoside,
        fluoroquinolone, tetracycline, sulfonamide, amphenicol,
        macrolide, nitrofuran),
      ~ . == TRUE
    ), na.rm = TRUE),
    DR  = n_classes_resistant >= 1,
    MDR = n_classes_resistant >= 3
  )
```

# Overall DR and MDR estimates

```{r overall-dr-mdr}
# Overall proportion of DR and MDR isolates with Wilson confidence intervals.
overall_dr_mdr <- AMR %>%
  summarise(
    total = n(),
    DR_n = sum(DR, na.rm = TRUE),
    MDR_n = sum(MDR, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(DR_n, MDR_n), names_to = "indicator", values_to = "n") %>%
  mutate(
    indicator = recode(indicator, DR_n = "DR", MDR_n = "MDR"),
    percent = n / total * 100
  ) %>%
  rowwise() %>%
  mutate(
    ci = list(binom.confint(n, total, method = "wilson")),
    ci_low = ci$lower * 100,
    ci_high = ci$upper * 100
  ) %>%
  ungroup() %>%
  select(indicator, n, total, percent, ci_low, ci_high)

overall_dr_mdr
```

# Overall resistance by antibiotic

```{r overall-resistance}
# Calculate resistance proportion for each antibiotic.
resistance_rate <- AMR %>%
  summarise(across(all_of(ast_cols), ~ {
    total <- sum(!is.na(.))
    resistant <- sum(. == "R", na.rm = TRUE)
    tibble(resistant = resistant, tested = total)
  })) %>%
  pivot_longer(everything(), names_to = "Antibiotic", values_to = "data") %>%
  unnest_wider(data) %>%
  mutate(
    percent = resistant / tested * 100,
    resistance = paste0(resistant, "/", tested, " (", round(percent, 1), "%)")
  ) %>%
  arrange(desc(percent))

resistance_rate
```

```{r overall-resistance-plot, fig.width=7, fig.height=7}
# Bar plot of resistance percentage by antibiotic, including DR and MDR at the end.
plot_data_overall <- resistance_rate %>%
  select(Antibiotic, percent) %>%
  arrange(percent) %>%
  bind_rows(
    overall_dr_mdr %>%
      transmute(Antibiotic = indicator, percent)
  ) %>%
  mutate(Antibiotic = factor(Antibiotic, levels = Antibiotic))

ggplot(plot_data_overall, aes(x = Antibiotic, y = percent)) +
  geom_col(width = 0.7) +
  coord_flip() +
  geom_text(aes(label = sprintf("%.1f", percent)), hjust = -0.15, size = 3) +
  labs(x = NULL, y = "Resistance (%)") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 9),
    axis.title.x = element_text(face = "bold"),
    legend.position = "none"
  )
```

# Prepare long-format AMR data

```{r long-format}
# Convert wide AST dataset to long format for antibiotic-specific analyses.
AMR_long <- AMR %>%
  pivot_longer(
    cols = all_of(ast_cols),
    names_to = "Antibiotic",
    values_to = "Result"
  ) %>%
  mutate(
    Class = case_when(
      Antibiotic %in% c("AMP", "PIP", "TZP", "AMC", "SAM") ~ "Beta-lactam",
      Antibiotic %in% c("CTX", "CRO", "CAZ", "FEP", "CXM", "CFP") ~ "Cephalosporin",
      Antibiotic %in% c("IPM", "MEM", "ETP") ~ "Carbapenem",
      Antibiotic %in% c("AMK", "GEN", "NET") ~ "Aminoglycoside",
      Antibiotic %in% c("CIP", "NOR", "PEF") ~ "Fluoroquinolone",
      Antibiotic == "TCY" ~ "Tetracycline",
      Antibiotic == "SXT" ~ "Sulfonamide",
      Antibiotic == "CHL" ~ "Amphenicol",
      Antibiotic == "AZM" ~ "Macrolide",
      Antibiotic == "NIT" ~ "Nitrofuran",
      TRUE ~ "Other"
    )
  )
```

# Temporal trend in resistance by antibiotic

```{r antibiotic-year-summary}
# Estimate yearly resistance by antibiotic.
trend_data <- AMR_long %>%
  group_by(year, Antibiotic, Class) %>%
  summarise(
    tested = sum(!is.na(Result)),
    resistant = sum(Result == "R", na.rm = TRUE),
    resistance_rate = resistant / tested * 100,
    .groups = "drop"
  )
```

```{r antibiotic-year-heatmap, fig.width=8, fig.height=7}
# Heatmap of yearly resistance percentage by antibiotic.
trend_data %>%
  filter(!is.na(year), !is.na(Class), Class != "Other") %>%
  group_by(Antibiotic) %>%
  mutate(mean_resistance = mean(resistance_rate, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(mean_resistance) %>%
  mutate(
    Antibiotic = factor(Antibiotic, levels = unique(Antibiotic)),
    resistance_plot = ifelse(tested < 1, NA, resistance_rate)
  ) %>%
  ggplot(aes(x = year, y = Antibiotic, fill = resistance_plot)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradientn(
    colours = c("#fff5f0", "#fcbba1", "#fb6a4a", "#cb181d", "#67000d"),
    na.value = "grey92",
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    name = "Resistance (%)"
  ) +
  scale_x_continuous(breaks = 2018:2024) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 11),
    legend.title = element_text(face = "bold"),
    panel.spacing = unit(0.3, "lines")
  )
```

# Temporal trend in DR and MDR

```{r dr-mdr-year-summary}
# Annual DR and MDR proportion.
dr_mdr_trend <- AMR %>%
  filter(!is.na(year), !is.na(DR), !is.na(MDR)) %>%
  group_by(year) %>%
  summarise(
    total = n(),
    DR_n = sum(DR, na.rm = TRUE),
    MDR_n = sum(MDR, na.rm = TRUE),
    DR_pct = DR_n / total * 100,
    MDR_pct = MDR_n / total * 100,
    .groups = "drop"
  )

dr_mdr_trend
```

```{r dr-mdr-logistic-trend}
# Logistic regression for temporal trend in DR and MDR.
# The odds ratio represents the annual change in odds.
glm_DR <- glm(DR ~ year, data = AMR, family = binomial)
glm_MDR <- glm(MDR ~ year, data = AMR, family = binomial)

trend_models <- bind_rows(
  tidy(glm_DR, conf.int = TRUE, exponentiate = TRUE) %>% mutate(outcome = "DR"),
  tidy(glm_MDR, conf.int = TRUE, exponentiate = TRUE) %>% mutate(outcome = "MDR")
) %>%
  filter(term == "year") %>%
  select(outcome, estimate, conf.low, conf.high, p.value)

trend_models
```

# Serovar-specific AMR patterns

```{r serovar-amr-summary}
# Summarise DR and MDR by serovar.
serovar_amr <- AMR %>%
  filter(!is.na(serotype), serotype != "Unavailable") %>%
  group_by(serotype) %>%
  summarise(
    total = n(),
    DR_n = sum(DR, na.rm = TRUE),
    MDR_n = sum(MDR, na.rm = TRUE),
    DR_pct = DR_n / total * 100,
    MDR_pct = MDR_n / total * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(total))

serovar_amr
```

```{r selected-serovars}
# Define selected serovars for detailed analysis.
# These include dominant serovars and additional serovars of public health interest.
main_serovars <- c(
  "Salmonella Enteritidis",
  "Salmonella Weltevreden",
  "Salmonella stanley",
  "Salmonella Paratyphi B",
  "Salmonella Typhimurium",
  "Salmonella Newport",
  "Salmonella Agona",
  "Salmonella Bareilly",
  "Salmonella ser. Hindmarsh",
  "Salmonella Corvalis"
)

extra_serovars <- c(
  "Salmonella Albany",
  "Salmonella Brancaster",
  "Salmonella Gallinarum",
  "4,[5],12:i:-"
)

all_selected <- c(main_serovars, extra_serovars)
```

```{r serovar-mdr-trend}
# Assess annual trend in MDR for selected serovars using logistic regression.
serovar_trend <- AMR %>%
  filter(serotype %in% all_selected, !is.na(MDR), !is.na(year)) %>%
  group_by(serotype) %>%
  group_modify(~ {
    model <- glm(MDR ~ year, data = .x, family = binomial)
    coef_df <- tidy(model) %>% filter(term == "year")
    ci <- confint.default(model)["year", ]
    tibble(
      OR = exp(coef_df$estimate),
      OR_low = exp(ci[1]),
      OR_high = exp(ci[2]),
      p_value = coef_df$p.value
    )
  }) %>%
  ungroup() %>%
  mutate(
    trend = case_when(
      p_value < 0.05 & OR > 1 ~ "Increasing",
      p_value < 0.05 & OR < 1 ~ "Decreasing",
      TRUE ~ "Stable"
    )
  ) %>%
  arrange(p_value)

serovar_trend
```

```{r serovar-model-diagnostics}
# Simple model stability checks for serovar-specific MDR trend models.
diag_results <- AMR %>%
  filter(serotype %in% all_selected) %>%
  group_split(serotype) %>%
  map_df(function(df) {
    sero_name <- unique(df$serotype)
    df_clean <- df %>% filter(!is.na(MDR), !is.na(year))

    yearly_counts <- df_clean %>%
      group_by(year) %>%
      summarise(
        total_year = n(),
        MDR_cases = sum(MDR),
        .groups = "drop"
      )

    m0 <- glm(MDR ~ 1, data = df_clean, family = binomial)
    m1 <- glm(MDR ~ year, data = df_clean, family = binomial)
    coef_year <- tidy(m1) %>% filter(term == "year")
    lrt <- anova(m0, m1, test = "Chisq")

    tibble(
      serotype = sero_name,
      n = nrow(df_clean),
      OR = exp(coef_year$estimate),
      p_value = coef_year$p.value,
      LRT_p = lrt$`Pr(>Chi)`[2],
      dispersion = deviance(m1) / df.residual(m1),
      std_error = coef_year$std.error,
      min_year_n = min(yearly_counts$total_year, na.rm = TRUE),
      zero_mdr_years = sum(yearly_counts$MDR_cases == 0, na.rm = TRUE),
      all_mdr_years = sum(yearly_counts$MDR_cases == yearly_counts$total_year, na.rm = TRUE)
    )
  }) %>%
  mutate(
    trend = case_when(
      p_value < 0.05 & OR > 1 ~ "Increasing",
      p_value < 0.05 & OR < 1 ~ "Decreasing",
      TRUE ~ "Stable"
    ),
    stability_flag = case_when(
      std_error > 0.30 ~ "Caution",
      OR > 5 | OR < 0.20 ~ "Caution",
      min_year_n < 10 ~ "Caution",
      zero_mdr_years > 2 ~ "Caution",
      all_mdr_years > 0 ~ "Caution",
      TRUE ~ "OK"
    ),
    fit_flag = case_when(
      dispersion > 2 ~ "Poor fit",
      dispersion > 1.5 ~ "Check fit",
      TRUE ~ "Acceptable"
    )
  ) %>%
  arrange(p_value)

diag_results
```

# Geographical variation by state and region

```{r define-region}
# Define Malaysian regions for subnational AMR analysis.
AMR <- AMR %>%
  mutate(region = case_when(
    state %in% c("Selangor", "Wilayah Persekutuan") ~ "Central",
    state %in% c("Pulau Pinang", "Kedah", "Perlis", "Perak") ~ "Northern",
    state %in% c("Kelantan", "Terengganu", "Pahang") ~ "East Coast",
    state %in% c("Johor", "Melaka", "Negeri Sembilan") ~ "Southern",
    state == "Sabah" ~ "Sabah",
    state == "Sarawak" ~ "Sarawak",
    TRUE ~ NA_character_
  ))
```

```{r state-region-summary}
# Summarise DR and MDR by state and region.
state_summary <- AMR %>%
  filter(!is.na(state)) %>%
  group_by(region, state) %>%
  summarise(
    total = n(),
    DR_n = sum(DR, na.rm = TRUE),
    MDR_n = sum(MDR, na.rm = TRUE),
    DR_pct = DR_n / total * 100,
    MDR_pct = MDR_n / total * 100,
    .groups = "drop"
  ) %>%
  arrange(region, desc(MDR_pct))

state_summary
```

```{r state-mdr-trend}
# Assess annual MDR trend by state.
state_trend <- AMR %>%
  filter(!is.na(MDR), !is.na(year), !is.na(state)) %>%
  group_by(region, state) %>%
  group_modify(~ {
    model <- glm(MDR ~ year, data = .x, family = binomial)
    coef_df <- tidy(model) %>% filter(term == "year")
    ci <- confint.default(model)["year", ]
    tibble(
      OR = exp(coef_df$estimate),
      OR_low = exp(ci[1]),
      OR_high = exp(ci[2]),
      p_value = coef_df$p.value
    )
  }) %>%
  ungroup() %>%
  mutate(
    trend = case_when(
      p_value < 0.05 & OR > 1 ~ "Increasing",
      p_value < 0.05 & OR < 1 ~ "Decreasing",
      TRUE ~ "Stable"
    )
  )

state_trend
```

# MDR resistance combinations

```{r mdr-patterns}
# Identify antibiotic combinations among MDR isolates.
# The pattern lists antibiotics interpreted as resistant for each MDR isolate.
AMR_MDR <- AMR %>%
  filter(MDR == TRUE) %>%
  mutate(across(all_of(ast_cols), as.character)) %>%
  mutate(
    MDR_pattern = apply(
      select(., all_of(ast_cols)),
      1,
      function(x) paste(names(x)[!is.na(x) & x == "R"], collapse = "-")
    )
  )

# Most common MDR patterns overall
mdr_patterns_overall <- AMR_MDR %>%
  filter(MDR_pattern != "") %>%
  count(MDR_pattern, sort = TRUE) %>%
  mutate(percent = n / sum(n) * 100) %>%
  slice_head(n = 15)

mdr_patterns_overall
```

```{r mdr-patterns-region}
# Most common MDR patterns by region.
region_patterns <- AMR_MDR %>%
  filter(!is.na(region), MDR_pattern != "") %>%
  group_by(region, MDR_pattern) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(region) %>%
  mutate(percent = n / sum(n) * 100) %>%
  slice_max(n, n = 5) %>%
  ungroup()

region_patterns
```

# AMR by specimen type

```{r specimen-summary}
# Summarise DR and MDR by specimen type.
# This uses the variable 'Specimen'. If your dataset uses another name,
# update the variable accordingly.
specimen_summary <- AMR %>%
  filter(!is.na(Specimen)) %>%
  group_by(Specimen) %>%
  summarise(
    total = n(),
    DR_n = sum(DR, na.rm = TRUE),
    MDR_n = sum(MDR, na.rm = TRUE),
    DR_pct = DR_n / total * 100,
    MDR_pct = MDR_n / total * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(total))

specimen_summary
```

```{r specimen-dr-mdr-plot, fig.width=7, fig.height=5}
# Stacked bar plot of non-DR, DR-only and MDR by specimen type.
plot_data_specimen <- specimen_summary %>%
  mutate(
    non_DR = total - DR_n,
    DR_only = DR_n - MDR_n,
    MDR = MDR_n
  ) %>%
  select(Specimen, non_DR, DR_only, MDR) %>%
  pivot_longer(
    cols = c(non_DR, DR_only, MDR),
    names_to = "Category",
    values_to = "Count"
  ) %>%
  group_by(Specimen) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  ungroup() %>%
  mutate(
    Specimen = factor(Specimen, levels = specimen_summary %>% arrange(desc(total)) %>% pull(Specimen)),
    Category = recode(Category, non_DR = "Non-DR", DR_only = "DR", MDR = "MDR")
  )

ggplot(plot_data_specimen, aes(x = Specimen, y = Percent, fill = Category)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(
    data = specimen_summary,
    aes(x = Specimen, y = 105, label = paste0("n=", total)),
    inherit.aes = FALSE,
    size = 4
  ) +
  labs(x = NULL, y = "Percentage (%)") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_blank(),
    panel.grid = element_blank()
  )
```

```{r specimen-antibiotic-heatmap, fig.width=8, fig.height=8}
# Heatmap of antibiotic-specific resistance by year for blood and stool isolates.
# This uses the variable `Specimen type`. If your dataset uses another name,
# update the variable accordingly.
trend_data_specimen <- AMR_long %>%
  filter(`Specimen type` %in% c("blood", "stool")) %>%
  group_by(`Specimen type`, year, Antibiotic) %>%
  summarise(
    tested = sum(!is.na(Result)),
    resistant = sum(Result == "R", na.rm = TRUE),
    resistance_rate = resistant / tested * 100,
    .groups = "drop"
  )

trend_data_specimen %>%
  filter(!is.na(year)) %>%
  group_by(`Specimen type`, Antibiotic) %>%
  mutate(mean_resistance = mean(resistance_rate, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    Antibiotic = reorder_within(Antibiotic, mean_resistance, `Specimen type`),
    `Specimen type` = factor(`Specimen type`, levels = c("blood", "stool")),
    resistance_plot = ifelse(tested < 1, NA, resistance_rate)
  ) %>%
  ggplot(aes(x = year, y = Antibiotic, fill = resistance_plot)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradientn(
    colours = c("#fff5f0", "#fcbba1", "#fb6a4a", "#cb181d", "#67000d"),
    na.value = "grey92",
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    name = "Resistance (%)"
  ) +
  scale_x_continuous(breaks = 2018:2024) +
  facet_wrap(~ `Specimen type`, ncol = 1, scales = "free_y") +
  scale_y_reordered() +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    panel.spacing = unit(0.3, "lines")
  )
```

```{r dr-mdr-by-specimen-time, fig.width=7, fig.height=5}
# Annual DR and MDR trends overall and by selected specimen types.
overall <- AMR %>%
  filter(!is.na(year), !is.na(DR), !is.na(MDR)) %>%
  group_by(year) %>%
  summarise(
    DR = mean(DR) * 100,
    MDR = mean(MDR) * 100,
    .groups = "drop"
  ) %>%
  mutate(Group = "Overall")

by_specimen <- AMR %>%
  filter(`Specimen type` %in% c("blood", "stool"), !is.na(year), !is.na(DR), !is.na(MDR)) %>%
  group_by(`Specimen type`, year) %>%
  summarise(
    DR = mean(DR) * 100,
    MDR = mean(MDR) * 100,
    .groups = "drop"
  ) %>%
  rename(Group = `Specimen type`)

plot_data <- bind_rows(overall, by_specimen) %>%
  pivot_longer(cols = c(DR, MDR), names_to = "Type", values_to = "Percent") %>%
  mutate(Group = factor(Group, levels = c("Overall", "blood", "stool")))

ggplot(plot_data, aes(x = year, y = Percent, linetype = Group, shape = Group, color = Type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = 2018:2024) +
  labs(x = NULL, y = "Percentage (%)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    legend.position = "right"
  )
```

# AMR by age group

```{r age-summary}
# Summarise DR and MDR by age group.
AMR_age <- AMR %>%
  filter(!is.na(age_categ), age_categ != "Unknown")

age_summary <- AMR_age %>%
  group_by(age_categ) %>%
  summarise(
    total = n(),
    DR_n = sum(DR, na.rm = TRUE),
    MDR_n = sum(MDR, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    DR_pct = DR_n / total * 100,
    MDR_pct = MDR_n / total * 100,
    DR_ci_low = binom.confint(DR_n, total, method = "wilson")$lower * 100,
    DR_ci_high = binom.confint(DR_n, total, method = "wilson")$upper * 100,
    MDR_ci_low = binom.confint(MDR_n, total, method = "wilson")$lower * 100,
    MDR_ci_high = binom.confint(MDR_n, total, method = "wilson")$upper * 100
  ) %>%
  ungroup()

age_summary
```

```{r age-group-tests}
# Chi-square tests to compare DR and MDR distribution across age groups.
table_DR <- table(AMR_age$age_categ, AMR_age$DR)
table_MDR <- table(AMR_age$age_categ, AMR_age$MDR)

chisq.test(table_DR)
chisq.test(table_MDR)
```

```{r age-logistic-models}
# Logistic regression for association between age group and DR/MDR.
glm_DR_age <- glm(DR ~ age_categ, data = AMR_age, family = binomial)
glm_MDR_age <- glm(MDR ~ age_categ, data = AMR_age, family = binomial)

age_model_results <- bind_rows(
  tidy(glm_DR_age, conf.int = TRUE, exponentiate = TRUE) %>% mutate(outcome = "DR"),
  tidy(glm_MDR_age, conf.int = TRUE, exponentiate = TRUE) %>% mutate(outcome = "MDR")
)

age_model_results
```

```{r age-antibiotic-heatmap, fig.width=8, fig.height=8}
# Antibiotic-specific resistance trends for selected age groups.
# Update selected age groups according to the thesis analysis plan.
AMR_ce <- AMR_long %>%
  filter(age_categ %in% c("Children", "Elderly"))

trend_age_abx <- AMR_ce %>%
  group_by(age_categ, year, Antibiotic) %>%
  summarise(
    tested = sum(!is.na(Result)),
    resistant = sum(Result == "R", na.rm = TRUE),
    resistance_rate = resistant / tested * 100,
    .groups = "drop"
  )

trend_age_abx %>%
  filter(!is.na(year)) %>%
  group_by(age_categ, Antibiotic) %>%
  mutate(mean_resistance = mean(resistance_rate, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Antibiotic = reorder_within(Antibiotic, mean_resistance, age_categ)) %>%
  ggplot(aes(x = year, y = Antibiotic, fill = resistance_rate)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(
    colours = c("#fff5f0", "#fcbba1", "#fb6a4a", "#cb181d", "#67000d"),
    limits = c(0, 100),
    name = "Resistance (%)"
  ) +
  facet_wrap(~ age_categ, ncol = 1, scales = "free_y") +
  scale_y_reordered() +
  scale_x_continuous(breaks = 2018:2024) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank())
```

