---
title: "Multiplication Factor Analysis for Human Non-typhoidal Salmonella in Malaysia, 2018–2024"
author: "Hazlienor Mohd Hatta"
subtitle: "R script for DrPH Thesis: Incidence, Serovar Dynamics and Forecasting of Human Non-typhoidal Salmonella Disease in Malaysia"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This Quarto document contains the R workflow for the multiplication factor (MF) analysis conducted for the DrPH thesis **"Incidence, Serovar Dynamics and Forecasting of Human Non-typhoidal Salmonella Disease in Malaysia"** by **Hazlienor Mohd Hatta**.

The purpose of this analysis is to estimate the extent of underestimation of human non-typhoidal *Salmonella* (NTS) infection in Malaysia using a surveillance pyramid approach. The workflow derives annual and overall multiplication factors for 2018–2024 by incorporating uncertainty in reporting, laboratory confirmation, healthcare-seeking behaviour, and symptomatic infection. Monte Carlo simulation is used to generate uncertainty intervals for total MF, estimated community cases, under-ascertainment (UA), and under-reporting (UR). A sensitivity analysis is also conducted to assess the influence of assumptions around laboratory confirmation probability.

# Analysis workflow

The analysis follows these main steps:

1.  Load required R packages.
2.  Define annual surveillance pyramid inputs.
3.  Specify surveillance pyramid levels and parameter assumptions.
4.  Estimate annual and overall multiplication factors using Monte Carlo simulation.
5.  Decompose total underestimation into under-ascertainment and under-reporting components.
6.  Conduct sensitivity analysis for the laboratory confirmation component.
7.  Visualise and compare base and sensitivity model estimates.

# 1. Load packages

```{r setup}
library(tidyverse)
```

# 2. Define surveillance pyramid input data

The input data summarise the annual counts used to estimate probabilities at each level of the surveillance pyramid. The final row represents the overall estimate for the full study period (2018–2024).

```{r input-data}
pyramid_data <- data.frame(
  year = c("2018", "2019", "2020", "2021", "2022", "2023", "2024", "2018-2024"),
  reported = c(2942, 3388, 3359, 2670, 2995, 3826, 4848, 24028),
  not_reported = c(688, 1222, 922, 711, 837, 1197, 876, 6453),
  lab_seek = c(168, 165, 164, 241, 52, 122, 327, 1239),
  seek = c(2959, 3371, 3126, 1628, 2762, 3433, 5042, 22321),
  symptom = c(4882, 4599, 3828, 2470, 4803, 6497, 9010, 36089)
)

pyramid_data
```

# 3. Define surveillance pyramid levels

The surveillance pyramid was conceptualised as follows:

-   **Level 5:** Reported laboratory-confirmed cases in national laboratory-based surveillance.
-   **Level 4:** Laboratory-confirmed cases among healthcare-seeking cases.
-   **Level 3:** Symptomatic cases who sought healthcare.
-   **Level 2:** Symptomatic infections.
-   **Level 1:** All infections, including symptomatic and asymptomatic infections.

The total multiplication factor was calculated by multiplying the inverse probabilities between consecutive surveillance levels.

# 4. Base Monte Carlo function

This function estimates the total multiplication factor and the corresponding estimated community cases. Uncertainty is propagated using Monte Carlo simulation with beta distributions for probability parameters.

```{r base-mc-function}
run_mf_simulation <- function(reported,
                              not_reported,
                              lab_seek,
                              seek,
                              symptom,
                              n_sim = 10000) {
  
  # Probability that a laboratory-confirmed case is reported
  p_report <- rbeta(n_sim, reported, not_reported)
  
  # Probability of laboratory confirmation among healthcare-seeking cases
  p_lab_seek <- rbeta(
    n_sim,
    shape1 = lab_seek,
    shape2 = pmax(seek - lab_seek, 1)
  )
  
  # Truncate very low laboratory-confirmation probabilities to avoid extreme values
  p_lab_seek <- pmax(p_lab_seek, 0.02)
  
  # Probability of seeking healthcare among symptomatic cases
  p_seek <- rbeta(
    n_sim,
    shape1 = seek,
    shape2 = pmax(symptom - seek, 1)
  )
  
  # Probability of being symptomatic among all infections
  # Assumption: 95% symptomatic and 5% asymptomatic
  p_symptom <- rbeta(n_sim, 95, 5)
  
  # Total multiplication factor
  mf_total <- (1 / p_report) *
    (1 / p_lab_seek) *
    (1 / p_seek) *
    (1 / p_symptom)
  
  # Estimated community cases
  true_cases <- reported * mf_total
  
  c(
    MF_low = quantile(mf_total, 0.025),
    MF_med = quantile(mf_total, 0.5),
    MF_high = quantile(mf_total, 0.975),
    true_low = quantile(true_cases, 0.025),
    true_med = quantile(true_cases, 0.5),
    true_high = quantile(true_cases, 0.975)
  )
}
```

# 5. Estimate total multiplication factor by year

```{r total-mf-results}
results_total_mf <- pyramid_data %>%
  rowwise() %>%
  mutate(res = list(run_mf_simulation(reported, not_reported, lab_seek, seek, symptom))) %>%
  mutate(
    MF_low = res[1],
    MF_med = res[2],
    MF_high = res[3],
    true_low = res[4],
    true_med = res[5],
    true_high = res[6]
  ) %>%
  ungroup() %>%
  select(year, MF_low, MF_med, MF_high, true_low, true_med, true_high)

results_total_mf
```

# 6. Decompose multiplication factors into UA and UR

This function estimates the component multiplication factors and separates total underestimation into:

-   **Under-ascertainment (UA):** loss before or at healthcare-seeking level, represented by symptomatic and healthcare-seeking components.
-   **Under-reporting (UR):** loss within the healthcare and surveillance system, represented by laboratory confirmation and reporting components.

```{r component-mc-function}
run_mf_component_simulation <- function(reported,
                                        not_reported,
                                        lab_seek,
                                        seek,
                                        symptom,
                                        n_sim = 10000) {
  
  # Reporting probability
  p_report <- rbeta(n_sim, reported, not_reported)
  
  # Laboratory-confirmation probability among healthcare-seeking cases
  p_lab_seek <- rbeta(
    n_sim,
    shape1 = lab_seek,
    shape2 = pmax(seek - lab_seek, 1)
  )
  p_lab_seek <- pmax(p_lab_seek, 0.02)
  
  # Healthcare-seeking probability among symptomatic cases
  p_seek <- rbeta(
    n_sim,
    shape1 = seek,
    shape2 = pmax(symptom - seek, 1)
  )
  
  # Symptomatic probability among all infections
  p_symptom <- rbeta(n_sim, 95, 5)
  
  # Component multiplication factors
  MF_report <- 1 / p_report
  MF_lab <- 1 / p_lab_seek
  MF_seek <- 1 / p_seek
  MF_symptom <- 1 / p_symptom
  
  # Under-ascertainment and under-reporting components
  UA <- MF_symptom * MF_seek
  UR <- MF_lab * MF_report
  
  # Total multiplication factor and estimated community cases
  MF_total <- UA * UR
  true_cases <- reported * MF_total
  
  c(
    MF_low = quantile(MF_total, 0.025),
    MF_med = quantile(MF_total, 0.5),
    MF_high = quantile(MF_total, 0.975),
    true_low = quantile(true_cases, 0.025),
    true_med = quantile(true_cases, 0.5),
    true_high = quantile(true_cases, 0.975),
    MF_report = quantile(MF_report, 0.5),
    MF_lab = quantile(MF_lab, 0.5),
    MF_seek = quantile(MF_seek, 0.5),
    MF_symptom = quantile(MF_symptom, 0.5),
    UA_low = quantile(UA, 0.025),
    UA_med = quantile(UA, 0.5),
    UA_high = quantile(UA, 0.975),
    UR_low = quantile(UR, 0.025),
    UR_med = quantile(UR, 0.5),
    UR_high = quantile(UR, 0.975)
  )
}
```

```{r component-results}
results_components <- pyramid_data %>%
  rowwise() %>%
  mutate(res = list(run_mf_component_simulation(reported, not_reported, lab_seek, seek, symptom))) %>%
  mutate(
    MF_low = res[1],
    MF_med = res[2],
    MF_high = res[3],
    true_low = res[4],
    true_med = res[5],
    true_high = res[6],
    MF_report = res[7],
    MF_lab = res[8],
    MF_seek = res[9],
    MF_symptom = res[10],
    UA_low = res[11],
    UA_med = res[12],
    UA_high = res[13],
    UR_low = res[14],
    UR_med = res[15],
    UR_high = res[16]
  ) %>%
  ungroup() %>%
  select(
    year,
    MF_low, MF_med, MF_high,
    UA_low, UA_med, UA_high,
    UR_low, UR_med, UR_high,
    MF_report, MF_lab, MF_seek, MF_symptom,
    true_low, true_med, true_high
  )

results_components
```

# 7. Format uncertainty intervals

```{r format-results}
format_ui <- function(low, med, high, digits = 1) {
  paste0(round(med, digits), " (", round(low, digits), ", ", round(high, digits), ")")
}

results_components_formatted <- results_components %>%
  mutate(
    MF = format_ui(MF_low, MF_med, MF_high),
    UA = format_ui(UA_low, UA_med, UA_high),
    UR = format_ui(UR_low, UR_med, UR_high)
  ) %>%
  select(year, MF, UA, UR, MF_report, MF_lab, MF_seek, MF_symptom, true_low, true_med, true_high)

results_components_formatted
```

# 8. Sensitivity analysis for laboratory confirmation probability

The total multiplication factor is highly influenced by the probability of laboratory confirmation among healthcare-seeking cases. A sensitivity analysis was therefore conducted by replacing the outbreak-derived laboratory confirmation probability with a uniform assumption ranging from 3% to 10%.

```{r sensitivity-function}
run_mf_sensitivity <- function(reported,
                               not_reported,
                               lab_seek,
                               seek,
                               symptom,
                               n_sim = 10000,
                               mode = "base") {
  
  # Reporting probability
  p_report <- rbeta(n_sim, reported, not_reported)
  
  # Laboratory-confirmation probability
  if (mode == "base") {
    # Base model: outbreak-derived laboratory confirmation probability
    p_lab_seek <- rbeta(
      n_sim,
      shape1 = lab_seek,
      shape2 = pmax(seek - lab_seek, 1)
    )
  } else if (mode == "sensitivity") {
    # Sensitivity model: assumed 3%–10% laboratory confirmation probability
    p_lab_seek <- runif(n_sim, 0.03, 0.10)
  } else {
    stop("mode must be either 'base' or 'sensitivity'.")
  }
  
  p_lab_seek <- pmax(p_lab_seek, 0.02)
  
  # Healthcare-seeking probability
  p_seek <- rbeta(
    n_sim,
    shape1 = seek,
    shape2 = pmax(symptom - seek, 1)
  )
  
  # Symptomatic probability
  p_symptom <- rbeta(n_sim, 95, 5)
  
  # Total multiplication factor
  MF_total <- (1 / p_report) *
    (1 / p_lab_seek) *
    (1 / p_seek) *
    (1 / p_symptom)
  
  true_cases <- reported * MF_total
  
  c(
    MF_low = quantile(MF_total, 0.025),
    MF_med = quantile(MF_total, 0.5),
    MF_high = quantile(MF_total, 0.975),
    true_low = quantile(true_cases, 0.025),
    true_med = quantile(true_cases, 0.5),
    true_high = quantile(true_cases, 0.975)
  )
}
```

```{r base-vs-sensitivity-results}
results_base <- pyramid_data %>%
  rowwise() %>%
  mutate(res = list(run_mf_sensitivity(reported, not_reported, lab_seek, seek, symptom, mode = "base"))) %>%
  mutate(
    MF_low = res[1],
    MF_med = res[2],
    MF_high = res[3],
    true_low = res[4],
    true_med = res[5],
    true_high = res[6]
  ) %>%
  ungroup() %>%
  select(year, MF_low, MF_med, MF_high, true_low, true_med, true_high)

results_sensitivity <- pyramid_data %>%
  rowwise() %>%
  mutate(res = list(run_mf_sensitivity(reported, not_reported, lab_seek, seek, symptom, mode = "sensitivity"))) %>%
  mutate(
    MF_low_sens = res[1],
    MF_med_sens = res[2],
    MF_high_sens = res[3],
    true_low_sens = res[4],
    true_med_sens = res[5],
    true_high_sens = res[6]
  ) %>%
  ungroup() %>%
  select(year, MF_low_sens, MF_med_sens, MF_high_sens, true_low_sens, true_med_sens, true_high_sens)

final_compare <- left_join(results_base, results_sensitivity, by = "year")

final_compare
```

# 9. Visualise base and sensitivity model estimates

```{r prepare-plot-data}
plot_base <- final_compare %>%
  select(year, low = MF_low, med = MF_med, high = MF_high) %>%
  mutate(model = "Outbreak-derived")

plot_sensitivity <- final_compare %>%
  select(year, low = MF_low_sens, med = MF_med_sens, high = MF_high_sens) %>%
  mutate(model = "Sensitivity (3–10%)")

plot_data <- bind_rows(plot_base, plot_sensitivity) %>%
  mutate(
    year = factor(
      year,
      levels = c("2018", "2019", "2020", "2021", "2022", "2023", "2024", "2018-2024")
    )
  )
```

```{r mf-comparison-plot, fig.width=9, fig.height=5}
ggplot(plot_data, aes(x = year, y = med, group = model, color = model, fill = model)) +
  geom_ribbon(aes(ymin = low, ymax = high), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(
    "Outbreak-derived" = "#D55E00",
    "Sensitivity (3–10%)" = "#0072B2"
  )) +
  scale_fill_manual(values = c(
    "Outbreak-derived" = "#D55E00",
    "Sensitivity (3–10%)" = "#0072B2"
  )) +
  labs(
    title = "Estimated Multiplication Factor (MF) by Year",
    subtitle = "Comparison of outbreak-derived and sensitivity-adjusted models",
    x = "Year",
    y = "Multiplication Factor (MF)",
    color = NULL,
    fill = NULL
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top",
    legend.title = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11)
  )
```

# Notes

-   The analysis uses a surveillance pyramid and multiplication factor approach to estimate underestimation of NTS infection.
-   The beta distribution was used to represent uncertainty around probability parameters derived from count data.
-   The laboratory confirmation component was explored through sensitivity analysis because it strongly influences the total multiplication factor.
-   No individual-level or identifiable patient data are included in this script.
