---
title: "Serovar Analysis of Human Non-typhoidal Salmonella in Malaysia, 2018–2024"
author: "Hazlienor Mohd Hatta"
subtitle: "R script for DrPH Thesis: Incidence, Serovar Dynamics and Forecasting of Human Non-typhoidal Salmonella Disease in Malaysia"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
editor: visual
---

# Introduction

This Quarto document contains the annotated R workflow used for the serovar analysis of laboratory-confirmed human non-typhoidal *Salmonella* (NTS) isolates in Malaysia from 2018 to 2024. The analysis forms part of the DrPH thesis titled **"Incidence, Serovar Dynamics and Forecasting of Human Non-typhoidal Salmonella Disease in Malaysia"** by **Hazlienor Mohd Hatta**.

The workflow includes data preparation, national serovar ranking, assessment of temporal stability, state-level and age-group variation, and heatmap visualisation of dominant serovars. The analysis excludes isolates recorded as unknown serovar where relevant, and describes proportional serovar distribution across selected demographic and geographical strata.

# 1. Load packages

```{r setup, message=FALSE, warning=FALSE}
# Core data management and visualisation
library(tidyverse)
library(forcats)
library(readxl)
library(scales)

# Statistical association and effect size
library(rcompanion)
```

# 2. Define file paths and analysis parameters

```{r parameters}
# Input file
input_file <- "NTS_inc.xlsx"

# Analysis period
analysis_years <- 2018:2024

# Label used in the dataset for isolates without assigned serovar
unknown_serovar <- "S. enterica subsp. enterica (serovar unknown)"

# Number of dominant serovars to display in selected plots
n_top_serovars <- 10

# State ordering used for figures and tables
state_order <- c(
  "Wilayah Persekutuan", "Selangor",
  "Kedah", "Pulau Pinang", "Perak", "Perlis",
  "Johor", "Melaka", "Negeri Sembilan",
  "Kelantan", "Terengganu", "Pahang",
  "Sabah", "Sarawak"
)

# Age ordering used for figures and tables
age_order <- c("Children", "Adolescents", "Adults", "Elderly")
```

# 3. Import and prepare data

```{r import-data}
# Import the NTS dataset.
# Expected variables include Year, State, age_who, Sex, serovar, and Warganegara.
NTS <- read_excel(input_file)

# Select variables required for serovar analysis and standardise labels.
sero <- NTS %>%
  select(Year, State, age_who, Sex, serovar, Warganegara) %>%
  mutate(
    Year = as.integer(Year),
    State = as.character(State),
    serovar = as.character(serovar),
    age_who = recode(as.character(age_who), "Older adults" = "Elderly")
  ) %>%
  filter(Year %in% analysis_years)

# Quick data check
sero %>%
  summarise(
    total_records = n(),
    years = paste(range(Year, na.rm = TRUE), collapse = "-"),
    states = n_distinct(State),
    serovars = n_distinct(serovar)
  )
```

# 4. National serovar ranking by year

This section summarises the top ten serovars nationally for each year and assesses whether dominant serovars remain stable over time.

```{r national-ranking-manual}
# Manually defined annual top-ten serovar ranking based on national frequency tables.
# Rank 1 indicates the most frequent serovar for that year.
srank <- tibble(
  Year = rep(analysis_years, each = 10),
  Rank = rep(1:10, times = length(analysis_years)),
  Serovar = c(
    # 2018
    "Enteritidis", "Weltevreden", "Paratyphi B", "Stanley", "Typhimurium",
    "Newport", "Bareilly", "Agona", "Saintpaul", "Bovismorbificans",

    # 2019
    "Enteritidis", "Weltevreden", "Paratyphi B", "Stanley", "Newport",
    "Hindmarsh", "Bareilly", "Bovismorbificans", "Agona", "Typhimurium",

    # 2020
    "Enteritidis", "Weltevreden", "Paratyphi B", "Typhimurium", "Stanley",
    "Hindmarsh", "Agona", "Bareilly", "Newport", "Corvallis",

    # 2021
    "Enteritidis", "Weltevreden", "Paratyphi B", "Stanley", "Typhimurium",
    "Hindmarsh", "Bareilly", "Agona", "Newport", "Corvallis",

    # 2022
    "Enteritidis", "Weltevreden", "Paratyphi B", "Stanley", "Hindmarsh",
    "Corvallis", "Typhimurium", "Newport", "Agona", "Bareilly",

    # 2023
    "Enteritidis", "Weltevreden", "Stanley", "Paratyphi B", "Typhimurium",
    "Bareilly", "Hindmarsh", "Newport", "Albany", "Agona",

    # 2024
    "Enteritidis", "Weltevreden", "Stanley", "Paratyphi B", "Bareilly",
    "Hindmarsh", "Newport", "Brancaster", "Agona", "Typhimurium"
  )
)
```

```{r national-ranking-summary}
# Frequency of appearance among the annual top-ten list.
rank_frequency <- srank %>%
  count(Serovar, sort = TRUE)

rank_frequency

# Summary of rank stability across years.
summary_table <- srank %>%
  group_by(Serovar) %>%
  summarise(
    Frequency = n(),
    Mean_Rank = round(mean(Rank), 2),
    Top3_Count = sum(Rank <= 3),
    .groups = "drop"
  ) %>%
  mutate(
    Category = case_when(
      Mean_Rank <= 3 ~ "Highly dominant",
      Mean_Rank <= 6 ~ "Moderately dominant",
      TRUE ~ "Occasional"
    )
  ) %>%
  arrange(Mean_Rank)

summary_table
```

```{r national-rank-heatmap, fig.width=8, fig.height=5}
# Heatmap of annual ranking among top-ten serovars.
ggplot(srank, aes(x = factor(Year), y = Serovar, fill = Rank)) +
  geom_tile(color = NA) +
  scale_fill_gradientn(
    colours = c("#08306B", "#2171B5", "#6BAED6", "#C6DBEF", "#DEEBF7"),
    breaks = 1:10,
    labels = 1:10
  ) +
  scale_y_discrete(limits = rev) +
  labs(x = NULL, y = NULL, fill = "Rank") +
  theme_classic() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```

# 5. State-level serovar distribution

This section examines geographical variation in serovar distribution across Malaysian states. Unknown serovars are excluded from the proportion denominator for this descriptive comparison.

```{r top-serovars-state}
# Identify the top serovars nationally, excluding unknown serovar.
top_serovars <- sero %>%
  filter(!is.na(serovar), serovar != unknown_serovar) %>%
  count(serovar, sort = TRUE) %>%
  slice_head(n = n_top_serovars) %>%
  pull(serovar)

top_serovars
```

```{r state-serovar-table}
# Calculate state-level serovar proportions.
# Denominator: all known serovars within each state.
state_sero <- sero %>%
  filter(!is.na(serovar), serovar != unknown_serovar) %>%
  count(State, serovar, name = "Cases") %>%
  group_by(State) %>%
  mutate(Total_known_serovar = sum(Cases), Prop = Cases / Total_known_serovar) %>%
  ungroup() %>%
  filter(serovar %in% top_serovars)

# Order serovars by mean state-level proportion.
state_serovar_order <- state_sero %>%
  group_by(serovar) %>%
  summarise(mean_prop = mean(Prop, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_prop)) %>%
  pull(serovar)

state_sero <- state_sero %>%
  mutate(
    State = factor(State, levels = state_order),
    serovar = factor(serovar, levels = rev(state_serovar_order))
  )

state_sero
```

```{r state-serovar-heatmap, fig.width=10, fig.height=6}
# Heatmap of dominant serovar proportions by state.
ggplot(state_sero, aes(x = State, y = serovar, fill = Prop)) +
  geom_tile(color = NA) +
  scale_fill_gradientn(
    colours = c("#DEEBF7", "#C6DBEF", "#6BAED6", "#2171B5", "#08306B"),
    name = "Proportion",
    breaks = c(0, 0.2, 0.4, 0.6),
    limits = c(0, 0.6),
    oob = scales::squish
  ) +
  labs(x = NULL, y = NULL) +
  theme_classic() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )
```

```{r state-serovar-association}
# Statistical test of association between state and serovar distribution.
# For sparse data, interpret chi-square results cautiously.
state_serovar_table <- sero %>%
  filter(!is.na(State), !is.na(serovar), serovar != unknown_serovar) %>%
  count(State, serovar) %>%
  pivot_wider(names_from = serovar, values_from = n, values_fill = 0) %>%
  column_to_rownames("State") %>%
  as.matrix()

chisq_state_serovar <- chisq.test(state_serovar_table)
chisq_state_serovar

# Effect size using Cramer's V.
cramerV(state_serovar_table)
```

# 6. State-year serovar distribution

This section describes temporal variation in dominant serovars within each state.

```{r state-year-serovar-data}
# Calculate proportions by state and year.
# Denominator: all known serovars within each state-year combination.
state_year_sero_plot <- sero %>%
  filter(!is.na(serovar), serovar != unknown_serovar) %>%
  count(State, Year, serovar, name = "Cases") %>%
  complete(State, Year = analysis_years, serovar, fill = list(Cases = 0)) %>%
  group_by(State, Year) %>%
  mutate(
    Total_known_serovar = sum(Cases),
    Prop = if_else(Total_known_serovar > 0, Cases / Total_known_serovar, NA_real_)
  ) %>%
  ungroup() %>%
  filter(serovar %in% top_serovars)

# Order variables for plotting.
state_year_sero_plot <- state_year_sero_plot %>%
  mutate(
    Year = factor(Year, levels = analysis_years),
    State = factor(State, levels = state_order),
    serovar = factor(serovar, levels = rev(state_serovar_order))
  )
```

```{r state-year-serovar-heatmap, fig.width=12, fig.height=10}
# Faceted heatmap showing state-specific serovar patterns by year.
ggplot(state_year_sero_plot, aes(x = Year, y = serovar, fill = Prop)) +
  geom_tile(color = NA) +
  scale_fill_gradientn(
    colours = c("#DEEBF7", "#C6DBEF", "#6BAED6", "#2171B5", "#08306B"),
    name = "Proportion",
    breaks = c(0, 0.2, 0.4, 0.6),
    limits = c(0, 0.6),
    oob = scales::squish,
    na.value = "grey90"
  ) +
  facet_wrap(~ State, ncol = 3, drop = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_classic() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    strip.text = element_text(size = 11, face = "bold"),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )
```

# 7. Age-group serovar distribution

This section examines variation in serovar distribution across age groups and over time. Records with unknown age group are excluded from age-stratified analyses.

```{r age-serovar-data}
# Identify top serovars among records with known age group.
top_serovars_age <- sero %>%
  filter(!is.na(serovar), serovar != unknown_serovar, age_who != "Unknown") %>%
  count(serovar, sort = TRUE) %>%
  slice_head(n = n_top_serovars) %>%
  pull(serovar)

# Calculate age-year serovar proportions.
# Denominator: all known serovars within each age-year combination.
age_year_sero_plot <- sero %>%
  filter(!is.na(serovar), serovar != unknown_serovar, age_who != "Unknown") %>%
  count(age_who, Year, serovar, name = "Cases") %>%
  complete(age_who, Year = analysis_years, serovar, fill = list(Cases = 0)) %>%
  group_by(age_who, Year) %>%
  mutate(
    Total_known_serovar = sum(Cases),
    Prop = if_else(Total_known_serovar > 0, Cases / Total_known_serovar, NA_real_)
  ) %>%
  ungroup() %>%
  filter(serovar %in% top_serovars_age)

# Order serovars based on mean proportion across age groups and years.
age_serovar_order <- age_year_sero_plot %>%
  group_by(serovar) %>%
  summarise(mean_prop = mean(Prop, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_prop)) %>%
  pull(serovar)

age_year_sero_plot <- age_year_sero_plot %>%
  mutate(
    age_who = factor(age_who, levels = age_order),
    Year = factor(Year, levels = analysis_years),
    serovar = factor(serovar, levels = rev(age_serovar_order))
  )
```

```{r age-year-serovar-heatmap, fig.width=10, fig.height=7}
# Faceted heatmap showing age-specific serovar patterns by year.
ggplot(age_year_sero_plot, aes(x = Year, y = serovar, fill = Prop)) +
  geom_tile(color = NA) +
  scale_fill_gradientn(
    colours = c("#DEEBF7", "#C6DBEF", "#6BAED6", "#2171B5", "#08306B"),
    name = "Proportion",
    breaks = c(0, 0.2, 0.4, 0.6),
    limits = c(0, 0.6),
    oob = scales::squish,
    na.value = "grey90"
  ) +
  facet_wrap(~ age_who, ncol = 2, drop = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_classic() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11),
    strip.text = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )
```

```{r age-serovar-summary}
# Top ten serovars within each age group.
top_sero_age <- sero %>%
  filter(!is.na(serovar), serovar != unknown_serovar, age_who != "Unknown") %>%
  count(age_who, serovar, name = "Cases") %>%
  group_by(age_who) %>%
  slice_max(order_by = Cases, n = n_top_serovars, with_ties = FALSE) %>%
  arrange(age_who, desc(Cases)) %>%
  mutate(Rank = row_number()) %>%
  ungroup()

top_sero_age

# Distribution of records by age group.
age_distribution <- sero %>%
  count(age_who, name = "Cases") %>%
  mutate(
    Total = sum(Cases),
    Prop = Cases / Total,
    Percent = Prop * 100
  )

age_distribution
```

# Notes for interpretation

-   Unknown serovars were excluded from serovar proportion analyses to avoid diluting the distribution of assigned serovars.
-   Proportions were calculated within each stratum, such as state, state-year, or age-year, to describe relative serovar composition.
-   Chi-square tests evaluate statistical association between state and serovar distribution but should be interpreted cautiously when sparse cells are present.
-   Cramer's V provides an effect size for the association between state and serovar distribution.
-   Heatmaps are descriptive and intended to support interpretation of dominant and secondary serovar patterns across time, geography, and age group.
